---
title: "Getting started with infosigasp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with infosigasp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Network access and a ~120 MB download are required for the live examples, so
# code that hits the network is shown but not evaluated when the vignette is
# built. Set options(infosigasp.run_live = TRUE) to run them locally. The
# chunk option must be an expression that also evaluates during R CMD build's
# tangle step (where code chunks never run), hence getOption() rather than a
# variable defined here.
```

## Overview

[INFOSIGA-SP](https://infosiga.detran.sp.gov.br/) is the São Paulo State
Traffic Accident Information and Management System, maintained by DETRAN-SP.
It publishes, as open data, detailed records of every traffic crash in the
state of São Paulo from 2015 onward.

`infosigasp` downloads and imports those records. It handles what makes the
raw files awkward to read directly.

- The files use **Latin-1** (ISO-8859-1), not UTF-8.
- **Semicolons** (`;`) separate the fields.
- Decimal numbers, such as coordinates, use a **comma** decimal mark.
- Dates follow **`DD/MM/YYYY`**.
- Each dataset spans **two files** (2015--2021 and 2022 onward).

```{r setup}
library(infosigasp)
```

## The three datasets

INFOSIGA-SP organises its data into three linked tables.

```{r datasets}
infosiga_datasets()
```

- **`sinistros`** — crash *events*. One row per recorded event, with the
  date, time, location (including latitude/longitude), road attributes and a
  breakdown of how many vehicles and victims were involved, by type and
  severity.
- **`pessoas`** — *people* (victims). One row per person involved, with
  demographic attributes, injury severity and, for fatalities, the date and
  place of death.
- **`veiculos`** — *vehicles*. One row per vehicle involved, with make/model,
  manufacturing and model years, colour and type.

All three share the `id_sinistro` key, so they can be joined together;
`pessoas` and `veiculos` additionally share `id_veiculo`.

## Reading data

The main entry point is `read_infosiga()`. The first call downloads the
source archive (about 120 MB) into a per-user cache; subsequent calls read
from that cache, so you only pay the download cost once.

```{r read, eval = isTRUE(getOption("infosigasp.run_live"))}
sinistros <- read_infosiga("sinistros")
sinistros
```

The `year` argument restricts the import to specific years, filtering on the
year of the crash (`ano_sinistro`).

```{r read-year, eval = isTRUE(getOption("infosigasp.run_live"))}
recent <- read_infosiga("sinistros", year = 2022:2025)
```

### Processed vs. raw data

By default `read_infosiga()` returns a **processed** dataset (`clean = TRUE`).
The processing never renames columns, recodes category labels or drops rows. It
only fixes types and source artefacts. The full, ordered list of steps lives in
`?clean_infosiga`; the main ones are below.

- **Dates and times.** Full dates are parsed to `Date` and times to `hms` (in
  both modes), and the `ano_mes_*` year-month columns (published as `"YYYY/MM"`)
  become first-of-month `Date` values.
- **Whitespace.** Text columns lose their leading and trailing whitespace. The
  source pads some fields to a fixed width (`nacionalidade` ships as
  `"BRASILEIRA          "`); untrimmed, those values break grouping and joins.
- **Missing values.** The `"NAO DISPONIVEL"` ("not available") marker becomes
  `NA` (trimming runs first, so space-padded markers are caught).
- **Ordered factors.** Ordinal columns sort and plot in their natural order
  instead of alphabetically.
    - `dia_da_semana`: `Domingo` < ... < `Sábado` (the Brazilian week starts on
      Sunday);
    - `turno`: `MADRUGADA` < `MANHA` < `TARDE` < `NOITE`;
    - `gravidade_lesao` (victims): `LEVE` < `GRAVE` < `FATAL`;
    - `faixa_etaria_demografica` / `faixa_etaria_legal`: age bands in order.
- **Crash-type flags.** The binary `tp_sinistro_*` columns (`"S"` / empty)
  become **logical**, so you can `sum()` or filter them directly. The
  categorical `tp_sinistro_primario` stays text.
- **Numeric strings.** `tempo_sinistro_obito` (days from crash to death) becomes
  **integer**, and `numero_logradouro` loses the spurious trailing `".0"` the
  export appends (`"193.0"`).
- **Coordinates.** `latitude`/`longitude` outside the São Paulo state bounding
  box become `NA` as a pair. These are mis-encoded values and `(0, 0)`
  placeholders, a few percent of records. No rows are dropped; pass
  `clean = FALSE` for the raw coordinates.

```{r clean, eval = isTRUE(getOption("infosigasp.run_live"))}
sinistros <- read_infosiga("sinistros")
levels(sinistros$dia_da_semana)
```

`dia_da_semana` is an ordered factor, so a weekday tabulation comes out in
calendar order rather than alphabetically.

```{r wday, eval = isTRUE(getOption("infosigasp.run_live"))}
table(sinistros$dia_da_semana)
```

Pass `clean = FALSE` if you would rather have the data exactly as published,
with every text column a character vector and `"NAO DISPONIVEL"` and the
source's whitespace padding preserved verbatim.

```{r raw, eval = isTRUE(getOption("infosigasp.run_live"))}
raw <- read_infosiga("sinistros", clean = FALSE)
```

`clean_infosiga()` applies the same processing to a raw import after the fact.

```{r clean-fn, eval = isTRUE(getOption("infosigasp.run_live"))}
processed <- clean_infosiga(raw, "sinistros")
```

### Tidying the category labels

`clean_infosiga()` fixes types and source artefacts but never touches category
*labels*, so that anyone reproducing an official DETRAN-SP figure gets exactly
the categories DETRAN-SP publishes. Those labels are, however, genuinely messy.
For your own analysis, run the opt-in second pass `tidy_infosiga_labels()`.

```{r tidy-labels, eval = isTRUE(getOption("infosigasp.run_live"))}
veiculos <- read_infosiga("veiculos") |>
  tidy_infosiga_labels("veiculos")

sort(table(veiculos$cor_veiculo), decreasing = TRUE)
```

Vehicle colour is the clearest case. The source carries dozens of distinct
values for **about sixteen real colours**, because two upstream systems coexist
in every year, one upper-case and one title-cased, and the two disagree on
gender agreement.

| source values | tidied |
|---------------|--------|
| `PRETA`, `Preta` | `Preta` |
| `BRANCA`, `Branco`, `BRANCA (PADRAO PM)` | `Branca` |
| `VERMELHA`, `Vermelho` | `Vermelha` |
| `CIN/VER/PRE`, `BRANCA/VERMELHA` | `Multicor` |

Note that `toupper()` alone will not merge `Branca` and `Branco`.

The remaining changes are listed below.

- **Place names.** The source publishes `municipio` unaccented (`"SAO PAULO"`)
  while `regiao_administrativa` keeps its accents, so neither joins to other
  Brazilian data as published. Both take the official IBGE spelling.
- **`profissao`** is Title Cased, which merges the several hundred values that
  differ from another only by capitalisation, and the stray `"NAO INFORMADA"`
  markers become `NA`.
- **`conservacao`** mixes the name of the maintaining body (`"PREFEITURA"`)
  with bare route codes (`"10.03"`); the codes move to a new
  `conservacao_codigo` column.

Two limits are worth naming. `profissao` keeps its unaccented spelling and its
occupations stay ungrouped, because no authoritative list covers these free-text
values. `conservacao` names keep their source capitalisation, because most are
brand acronyms (`SPMAR`, `TEBE`, `CART`) that Title Case would corrupt into
`Spmar` and `Tebe`.

### Joining to other Brazilian data

The `infosiga_municipios` lookup ships with the package and maps `cod_ibge` to
both spellings of each of the 645 municipalities.

```{r municipios}
head(infosiga_municipios, 4)
```

Always join on `cod_ibge`, never on the name. IBGE and INFOSIGA-SP spell nine
municipalities differently. Eight contain an apostrophe that INFOSIGA-SP renders
as a space (`"Santa Bárbara d'Oeste"` against `"SANTA BARBARA D OESTE"`), and
IBGE's `"São Luiz do Paraitinga"` appears as `"SAO LUIS DO PARAITINGA"`. A name
join loses all nine silently. `cod_ibge` is also the standard Brazilian
municipality key, so it joins INFOSIGA-SP to census and population data. Those
are the denominators that turn crash counts into crash *rates*.

### Inspecting the structure without downloading

The package ships a small sample of each dataset so you can inspect the
columns without any network access.

```{r sample}
sample_path <- system.file("extdata", "sinistros_sample.csv", package = "infosigasp")
sample <- readr::read_delim(
  sample_path,
  delim = ";",
  show_col_types = FALSE
)
dim(sample)
names(sample)
```

## Coverage and caveats

INFOSIGA-SP publishes one continuous series from 2015 onward, but what the
series *contains* has changed over time. These are properties of the source
data, not import artefacts, so `infosigasp` reports them rather than correcting
them. Several will quietly invalidate an analysis that looks perfectly
reasonable. `?read_infosiga` carries the full list; the two below catch people
out most often.

### 2015--2018 covers fatal crashes only

Non-fatal records begin in **2019**. Before that, every row is a fatal crash.
Counting crashes per year over the whole series therefore produces a roughly
20-fold jump in 2019 that reflects only the expansion of data collection.

| year | fatal crashes | non-fatal crashes |
|------|---------------|-------------------|
| 2015 | 5,942         | 0                 |
| 2018 | 4,869         | 0                 |
| 2019 | 4,804         | 116,412           |
| 2023 | 5,166         | 135,329           |

Those counts come from a 2026 release and shift as DETRAN-SP revises the data.
The break in 2019 does not move. The same break appears in `pessoas`, where the
`LEVE` and `GRAVE` injury levels simply do not occur before 2019.

So for any trend that reaches back before 2019, **restrict to fatalities**.

```{r fatal-only, eval = isTRUE(getOption("infosigasp.run_live"))}
library(dplyr)

read_infosiga("sinistros") |>
  filter(tipo_registro == "SINISTRO FATAL") |>
  count(ano_sinistro)
```

Otherwise, start the series in 2019 by passing a `year` range to
`read_infosiga()`.

### A third of `sinistros` rows are notifications

`tipo_registro` distinguishes confirmed crashes from `"NOTIFICACAO"`, a
reported event not yet confirmed as a crash. Notifications are about a third of
all rows, so treating every row as a crash overstates the total considerably.

```{r tipo-registro, eval = isTRUE(getOption("infosigasp.run_live"))}
read_infosiga("sinistros") |>
  count(tipo_registro, sort = TRUE)
```

DETRAN-SP reclassifies records as it validates them, so the newest months carry
an unusually high share of notifications. The last month or two of any release
is also incomplete. Drop the tail of the series before reading anything into a
recent trend.

### Smaller caveats

- **`tempo_sinistro_obito` is capped at 30 days**, the standard convention for
  attributing a death to a crash. Fatality counts here are 30-day counts.
- **The `qtd_*` vehicle columns disagree with the `veiculos` row count** for a
  small share of crashes. Pick one definition and say which. (The
  `qtd_gravidade_*` columns *do* match `pessoas` exactly.)
- **Coordinate availability varies by year.** It is highest through the middle
  of the series and lower at both ends, so a mapped subset is not a uniform
  time sample.
- **About a third of crashes have no `pessoas` row.** Use a left join to keep
  them.

## A short analysis

The imported data are ordinary tibbles, so any tidyverse (or base R) workflow
applies. Here is a count of traffic fatalities per year from the victims
dataset.

```{r fatalities, eval = isTRUE(getOption("infosigasp.run_live"))}
library(dplyr)

deaths_by_year <- read_infosiga("pessoas") |>
  filter(gravidade_lesao == "FATAL") |>
  count(ano_obito, name = "deaths") |>
  arrange(ano_obito)

deaths_by_year
```

That series is safe to read across all years because it counts only fatalities,
the one thing collected consistently since 2015.

Fatalities also break down by the type of victim (driver, passenger,
pedestrian).

```{r by-victim, eval = isTRUE(getOption("infosigasp.run_live"))}
read_infosiga("pessoas") |>
  filter(gravidade_lesao == "FATAL") |>
  count(tipo_de_vitima, sort = TRUE)
```

`sinistros` carries latitude and longitude as numeric columns, so you can map
crash locations directly or aggregate them by municipality
(`municipio` / `cod_ibge`).

## Managing the cache

The download lives in an operating-system specific cache directory.

```{r cache}
infosiga_cache_dir()
infosiga_cache_list()
```

DETRAN-SP refreshes the archive monthly. `infosiga_check_update()` compares
your cached copy against the mirror the package publishes, which a weekly job
re-fetches from DETRAN-SP, so it answers whether a newer archive exists without
downloading one.

```{r check-update, eval = isTRUE(getOption("infosigasp.run_live"))}
infosiga_check_update()
```

Force a re-download to pull the latest version.

```{r refresh, eval = isTRUE(getOption("infosigasp.run_live"))}
infosiga_download(overwrite = TRUE)
```

Clear the cache to reclaim disk space.

```{r clear, eval = isTRUE(getOption("infosigasp.run_live"))}
infosiga_cache_clear()
```

You can point the cache somewhere else for a session (or permanently via your
`.Rprofile`) with the `infosigasp.cache_dir` option.

```{r cache-opt, eval = FALSE}
options(infosigasp.cache_dir = "~/data/infosiga")
```

## The official data dictionary

INFOSIGA-SP distributes a field-by-field data dictionary (PDF, in Portuguese).
`infosiga_dictionary()` downloads it and returns the paths to the extracted
files.

```{r dictionary, eval = isTRUE(getOption("infosigasp.run_live"))}
pdfs <- infosiga_dictionary()
basename(pdfs)
```

## Citing the data

DETRAN-SP publishes the data under a Creative Commons Attribution 4.0 licence.
When you publish results based on these data, please cite INFOSIGA-SP /
DETRAN-SP as the source, <https://infosiga.detran.sp.gov.br/>.
