---
title: "Getting started with infosigasp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with infosigasp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Network access and a ~120 MB download are required for the live examples, so
# code that hits the network is shown but not evaluated when the vignette is
# built. Set this to TRUE to run them locally.
run_live <- FALSE
```

## Overview

[INFOSIGA-SP](https://infosiga.detran.sp.gov.br/) is the São Paulo State
Traffic Accident Information and Management System, maintained by DETRAN-SP.
It publishes, as open data, detailed records of every traffic crash in the
state of São Paulo from 2015 onward.

The `infosigasp` package wraps the download and import of those records. It
takes care of the things that make the raw files awkward to read directly:

- the files are encoded in **Latin-1** (ISO-8859-1), not UTF-8;
- fields are separated by **semicolons** (`;`);
- decimal numbers (such as coordinates) use a **comma** decimal mark;
- dates are formatted **`DD/MM/YYYY`**;
- each dataset is split across **two files** (2015--2021 and 2022 onward).

```{r setup}
library(infosigasp)
```

## The three datasets

INFOSIGA-SP organises its data into three linked tables:

```{r datasets}
infosiga_datasets()
```

- **`sinistros`** — crash *events*. One row per recorded event, with the
  date, time, location (including latitude/longitude), road attributes and a
  breakdown of how many vehicles and victims were involved, by type and
  severity.
- **`pessoas`** — *people* (victims). One row per person involved, with
  demographic attributes, injury severity and, for fatalities, the date and
  place of death.
- **`veiculos`** — *vehicles*. One row per vehicle involved, with make/model,
  manufacturing and model years, colour and type.

All three share the `id_sinistro` key, so they can be joined together;
`pessoas` and `veiculos` additionally share `id_veiculo`.

## Reading data

The main entry point is `read_infosiga()`. The first call downloads the
source archive (about 120 MB) into a per-user cache; subsequent calls read
from that cache, so you only pay the download cost once.

```{r read, eval = run_live}
sinistros <- read_infosiga("sinistros")
sinistros
```

You can restrict the import to specific years with the `year` argument, which
filters on the year of the crash (`ano_sinistro`):

```{r read-year, eval = run_live}
recent <- read_infosiga("sinistros", year = 2022:2025)
```

### A peek at the structure without downloading

The package ships a small sample of each dataset so you can inspect the
columns without any network access:

```{r sample}
sample_path <- system.file("extdata", "sinistros_sample.csv", package = "infosigasp")
sample <- readr::read_delim(
  sample_path,
  delim = ";",
  show_col_types = FALSE
)
dim(sample)
names(sample)
```

## A short analysis

Once imported, the data are ordinary tibbles, so any tidyverse (or base R)
workflow applies. For example, counting traffic fatalities per year from the
victims dataset:

```{r fatalities, eval = run_live}
library(dplyr)

deaths_by_year <- read_infosiga("pessoas") |>
  filter(gravidade_lesao == "FATAL") |>
  count(ano_obito, name = "deaths") |>
  arrange(ano_obito)

deaths_by_year
```

Or fatalities broken down by the type of victim (driver, passenger,
pedestrian):

```{r by-victim, eval = run_live}
read_infosiga("pessoas") |>
  filter(gravidade_lesao == "FATAL") |>
  count(tipo_de_vitima, sort = TRUE)
```

Because `sinistros` carries latitude and longitude as numeric columns, crash
locations can be mapped directly or aggregated by municipality
(`municipio` / `cod_ibge`).

## Managing the cache

The download lives in an operating-system specific cache directory:

```{r cache}
infosiga_cache_dir()
infosiga_cache_list()
```

The archive is refreshed monthly by DETRAN-SP. To pull the latest version,
force a re-download:

```{r refresh, eval = run_live}
infosiga_download(overwrite = TRUE)
```

To reclaim disk space, clear the cache:

```{r clear, eval = run_live}
infosiga_cache_clear()
```

You can point the cache somewhere else for a session (or permanently via your
`.Rprofile`) with the `infosigasp.cache_dir` option:

```{r cache-opt, eval = FALSE}
options(infosigasp.cache_dir = "~/data/infosiga")
```

## The official data dictionary

INFOSIGA-SP distributes a field-by-field data dictionary (PDF, in
Portuguese). `infosiga_dictionary()` downloads it and returns the paths to the
extracted files:

```{r dictionary, eval = run_live}
pdfs <- infosiga_dictionary()
basename(pdfs)
```

## Citing the data

Data are published by DETRAN-SP under a Creative Commons Attribution 4.0
licence. When you publish results based on these data, please cite
INFOSIGA-SP / DETRAN-SP as the source:
<https://infosiga.detran.sp.gov.br/>.
