---
title: "Getting started with infosigasp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with infosigasp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Network access and a ~120 MB download are required for the live examples, so
# code that hits the network is shown but not evaluated when the vignette is
# built. Set options(infosigasp.run_live = TRUE) to run them locally. The
# chunk option must be an expression that also evaluates during R CMD build's
# tangle step (where code chunks never run), hence getOption() rather than a
# variable defined here.
```

## Overview

[INFOSIGA-SP](https://infosiga.detran.sp.gov.br/) is the São Paulo State
Traffic Incident Information and Management System, maintained by DETRAN-SP.
It publishes open occurrence records from 2015 onward. Coverage and definitions
vary over time: 2015--2018 covers fatal crashes only, and the events table also
contains notifications that have not yet been confirmed as crashes.

`infosigasp` downloads and imports those records. It handles what makes the
raw files awkward to read directly.

- The files use **Latin-1** (ISO-8859-1), not UTF-8.
- **Semicolons** (`;`) separate the fields.
- Decimal numbers, such as coordinates, use a **comma** decimal mark.
- Dates follow **`DD/MM/YYYY`**.
- Each dataset spans **two files** (2015--2021 and 2022 onward).

```{r setup}
library(infosigasp)
```

## The three datasets

INFOSIGA-SP organises its data into three linked tables.

- **`sinistros`** — occurrence records. One row per confirmed crash or
  notification, with the date, time, location (including latitude/longitude),
  road attributes and a breakdown of how many vehicles and victims were
  involved, by type and severity.
- **`pessoas`** — *people* (victims). One row per person involved, with
  demographic attributes, injury severity and, for fatalities, the date and
  place of death.
- **`veiculos`** — *vehicles*. One row per vehicle involved, with make/model,
  manufacturing and model years, colour and type.

All three share the `id_sinistro` key, so they can be joined together;
`pessoas` and `veiculos` additionally share `id_veiculo`.

## Reading data

The main entry point is `read_infosiga()`. The first call downloads the
source archive (about 120 MB) into a per-user cache; subsequent calls read
from that cache, so you only pay the download cost once. In interactive use,
the package asks for confirmation before that first download.

```{r read, eval = isTRUE(getOption("infosigasp.run_live"))}
sinistros <- read_infosiga("sinistros")
sinistros
```

### Raw, typed and clean data

`read_infosiga()` separates tabular import, type parsing and cleaning through
the `processing` argument.

| mode | result |
|------|--------|
| `"raw"` | Every field is character. Empty strings, whitespace, sentinels and malformed representations remain visible. |
| `"typed"` | Documented dates, times and numeric fields receive R classes; other source values remain unchanged. |
| `"clean"` | The default. Applies the cleaning pipeline after typed import. |

Every mode decodes Latin-1 to UTF-8, parses the CSV structure and combines the
period files. Raw mode is therefore a lossless tabular representation rather
than a byte-for-byte copy of the source archive.

Choose the mode according to the task: use `"raw"` to audit source values,
`"typed"` to parse documented classes while retaining source labels and
padding, and the default `"clean"` mode for most analysis. All modes retain
import diagnostics in `attr(x, "problems")`; typed and clean modes replace
values that fail their declared parser with `NA`, while raw mode preserves the
field text.

```{r processing, eval = isTRUE(getOption("infosigasp.run_live"))}
raw <- read_infosiga("sinistros", processing = "raw")
typed <- read_infosiga("sinistros", processing = "typed")
clean <- read_infosiga("sinistros", processing = "clean")
```

Typed mode parses full dates to `Date`, times to `hms`, numeric fields to
integer or double, and empty fields to `NA`. Identifiers remain character to
preserve leading zeros and avoid integer overflow.

Clean mode then applies the following transformations:

- **Year-month dates.** The `ano_mes_*` columns (published as `"YYYY/MM"`)
  become first-of-month `Date` values.
- **Whitespace.** Text columns lose their leading and trailing whitespace. The
  source pads some fields to a fixed width (`nacionalidade` ships as
  `"BRASILEIRA          "`); untrimmed, those values break grouping and joins.
- **Missing values.** The `"NAO DISPONIVEL"` ("not available") marker becomes
  `NA` (trimming runs first, so space-padded markers are caught).
- **Ordered factors.** Ordinal columns sort and plot in their natural order
  instead of alphabetically.
    - `dia_da_semana`: `Domingo` < ... < `Sábado` (the Brazilian week starts on
      Sunday);
    - `turno`: `MADRUGADA` < `MANHA` < `TARDE` < `NOITE`;
    - `gravidade_lesao` (victims): `LEVE` < `GRAVE` < `FATAL`;
    - `faixa_etaria_demografica` / `faixa_etaria_legal`: age bands in order.
- **Crash-type flags.** The binary `tp_sinistro_*` columns (`"S"` / empty)
  become **logical**, so you can `sum()` or filter them directly. The
  categorical `tp_sinistro_primario` stays text.
- **Counts.** Missing `qtd_*` entries remain `NA`, including when another
  vehicle or injury-severity count is available for the same crash. Clean mode
  does not infer that an unreported count is zero.
- **Numeric strings.** `tempo_sinistro_obito` (days from crash to death) becomes
  **integer**, and `numero_logradouro` loses the spurious trailing `".0"` the
  export appends (`"193.0"`).
- **Coordinates.** Coordinate pairs outside the São Paulo state boundary with
  a 2 km buffer become `NA`. The buffer retains plausible near-border records,
  while the spatial test rejects extreme malformed values, `(0, 0)`
  placeholders and points elsewhere inside the state's rectangular extent. No
  rows are dropped; typed and raw modes preserve the coordinates before this
  validation.

Before converting an ordinal, flag or integer column, clean mode validates its
observed representations. If it finds an unexpected value, it preserves the
entire column and warns; the class can therefore remain character after an
upstream change.

These safeguards apply to closed-domain conversions, not to every parsed
column. For example, a malformed numeric coordinate can become `NA` during the
typed import and will be listed in `attr(x, "problems")`.

```{r clean, eval = isTRUE(getOption("infosigasp.run_live"))}
sinistros <- read_infosiga("sinistros")
levels(sinistros$dia_da_semana)
```

`dia_da_semana` is an ordered factor, so a weekday tabulation comes out in
calendar order rather than alphabetically.

```{r wday, eval = isTRUE(getOption("infosigasp.run_live"))}
table(sinistros$dia_da_semana)
```

### Standardizing category labels

Clean mode does not harmonize nominal categories beyond its documented
missing-value rule. Select only the harmonizations you need with `standardize`.

```{r standardize-labels, eval = isTRUE(getOption("infosigasp.run_live"))}
veiculos <- read_infosiga("veiculos", standardize = "cores")

sort(table(veiculos$cor_veiculo), decreasing = TRUE)
```

Vehicle colour is the clearest case. The source carries dozens of distinct
values for **about sixteen real colours**, because two upstream systems coexist
in every year, one upper-case and one title-cased, and the two disagree on
gender agreement.

| source values | standardized |
|---------------|--------------|
| `PRETA`, `Preta` | `Preta` |
| `BRANCA`, `Branco` | `Branca` |
| `VERMELHA`, `Vermelho` | `Vermelha` |

Note that `toupper()` alone will not merge `Branca` and `Branco`.
Detailed liveries and multi-tone values remain unchanged because grouping them
would be an analytical recode rather than label harmonization.

The remaining changes are listed below.

- **`"municipios"`.** Municipality names take their official IBGE spelling,
  matched by `cod_ibge`; INFOSIGA administrative-region names receive consistent
  accents and case without changing their classification.
- **`"profissoes"`.** Occupations are title-cased, which merges values that
  differ only by capitalisation. Known nonresponse markers—including variants
  of `"NAO INFORMADA"`, `"SEM INFORMACAO"` and `"DESCONHECIDO"`—become `NA`.

Use `standardize = c("municipios", "profissoes")` to combine options, or
`standardize = "all"` for every option applicable to the selected dataset.
`profissao` keeps its unaccented spelling and occupations stay ungrouped because
no authoritative list covers these free-text values. Standardization requires
`processing = "clean"`.

### Joining to other Brazilian data

Always join on `cod_ibge`, never on the name. IBGE and INFOSIGA-SP spell nine
municipalities differently. Eight contain an apostrophe that INFOSIGA-SP renders
as a space (`"Santa Bárbara d'Oeste"` against `"SANTA BARBARA D OESTE"`), and
IBGE's `"São Luiz do Paraitinga"` appears as `"SAO LUIS DO PARAITINGA"`. A name
join loses all nine silently. `cod_ibge` is also the standard Brazilian
municipality key, so it joins INFOSIGA-SP to census and population data. Those
are the denominators that turn crash counts into crash *rates*.

### Inspecting the structure without downloading

The package ships a small sample of each dataset so you can inspect the
columns without any network access.

```{r sample}
sample_path <- system.file("extdata", "sinistros_sample.csv", package = "infosigasp")
sample <- readr::read_delim(
  sample_path,
  delim = ";",
  show_col_types = FALSE
)
dim(sample)
names(sample)
```

## Coverage and caveats

INFOSIGA-SP publishes one continuous series from 2015 onward, but what the
series *contains* has changed over time. These are properties of the source
data, not import artefacts, so `infosigasp` reports them rather than correcting
them. Several will quietly invalidate an analysis that looks perfectly
reasonable. `?read_infosiga` carries the full list; the two below catch people
out most often.

### 2015--2018 covers fatal crashes only

Non-fatal records begin in **2019**. Before that, every row is a fatal crash.
Counting crashes per year over the whole series therefore produces a roughly
20-fold jump in 2019 that reflects only the expansion of data collection.

| year | fatal crashes | non-fatal crashes |
|------|---------------|-------------------|
| 2015 | 5,942         | 0                 |
| 2018 | 4,869         | 0                 |
| 2019 | 4,804         | 116,412           |
| 2023 | 5,166         | 135,329           |

Those counts come from a 2026 release and shift as DETRAN-SP revises the data.
The break in 2019 does not move. The same break appears in `pessoas`, where the
`LEVE` and `GRAVE` injury levels simply do not occur before 2019.

So for any trend that reaches back before 2019, **restrict to fatalities**.

```{r fatal-only, eval = isTRUE(getOption("infosigasp.run_live"))}
library(dplyr)

read_infosiga("sinistros") |>
  filter(tipo_registro == "SINISTRO FATAL") |>
  count(ano_sinistro)
```

Otherwise, start the series in 2019 by filtering `ano_sinistro >= 2019` after
importing the data.

### A third of `sinistros` rows are notifications

`tipo_registro` distinguishes confirmed crashes from `"NOTIFICACAO"`, a
reported event not yet confirmed as a crash. Notifications are about a third of
all rows, so treating every row as a crash overstates the total considerably.

```{r tipo-registro, eval = isTRUE(getOption("infosigasp.run_live"))}
read_infosiga("sinistros") |>
  count(tipo_registro, sort = TRUE)
```

DETRAN-SP reclassifies records as it validates them, so the newest months carry
an unusually high share of notifications. The last month or two of any release
is also incomplete. Drop the tail of the series before reading anything into a
recent trend.

### Smaller caveats

- **`tempo_sinistro_obito` is capped at 30 days**, the standard convention for
  attributing a death to a crash. Fatality counts here are 30-day counts.
- **The `qtd_*` vehicle columns disagree with the `veiculos` row count** for a
  small share of crashes. Pick one definition and say which. (The
  `qtd_gravidade_*` columns *do* match `pessoas` exactly.)
- **Coordinate availability varies by year.** It is highest through the middle
  of the series and lower at both ends, so a mapped subset is not a uniform
  time sample.
- **About a third of crashes have no `pessoas` row.** Use a left join to keep
  them.

## A short analysis

The imported data are ordinary tibbles, so any tidyverse (or base R) workflow
applies. Here is a count of traffic fatalities per year from the victims
dataset.

```{r fatalities, eval = isTRUE(getOption("infosigasp.run_live"))}
library(dplyr)

deaths_by_year <- read_infosiga("pessoas") |>
  filter(gravidade_lesao == "FATAL") |>
  count(ano_obito, name = "deaths") |>
  arrange(ano_obito)

deaths_by_year
```

That series is safe to read across all years because it counts only fatalities,
the one thing collected consistently since 2015.

Fatalities also break down by the type of victim (driver, passenger,
pedestrian).

```{r by-victim, eval = isTRUE(getOption("infosigasp.run_live"))}
read_infosiga("pessoas") |>
  filter(gravidade_lesao == "FATAL") |>
  count(tipo_de_vitima, sort = TRUE)
```

`sinistros` carries latitude and longitude as numeric columns, so you can map
crash locations directly or aggregate them by municipality
(`municipio` / `cod_ibge`).

## Updating the data

The package stores one local copy in the operating-system user cache. DETRAN-SP
updates the archive monthly. Use `refresh = TRUE` to replace your local copy
with the latest available version and return the requested dataset.

```{r refresh, eval = isTRUE(getOption("infosigasp.run_live"))}
sinistros <- read_infosiga("sinistros", refresh = TRUE)
```

`read_infosiga()` always imports every available year. Filter the returned
tibble when you need a shorter period.

```{r filter-years, eval = isTRUE(getOption("infosigasp.run_live"))}
recent <- read_infosiga("sinistros") |>
  filter(ano_sinistro >= 2022)
```

You can point the cache somewhere else for a session (or permanently via your
`.Rprofile`) with the `infosigasp.cache_dir` option.

```{r cache-opt, eval = FALSE}
options(infosigasp.cache_dir = "~/data/infosiga")
```

## The official data dictionary

INFOSIGA-SP distributes a field-by-field data dictionary (PDF, in Portuguese).
`dictionary_infosiga()` downloads it and returns the paths to the extracted
files. Pass `"sinistros"`, `"pessoas"` or `"veiculos"` to return one PDF.

```{r dictionary, eval = isTRUE(getOption("infosigasp.run_live"))}
pdfs <- dictionary_infosiga()
basename(pdfs)

sinistros_pdf <- dictionary_infosiga("sinistros")
```

## Citing the data

DETRAN-SP publishes the data under a Creative Commons Attribution 4.0 licence.
When you publish results based on these data, please cite INFOSIGA-SP /
DETRAN-SP as the source, <https://infosiga.detran.sp.gov.br/>.
