---
title: "Data dictionary"
css: data-dictionary.css
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

This article transcribes the official INFOSIGA-SP data dictionary
(**v1.5, 2026-06-16**), published by DETRAN-SP, for the three datasets
returned by `read_infosiga()`: `sinistros` (confirmed crashes and
notifications), `pessoas` (victims) and `veiculos` (vehicles). Field names,
descriptions and category codes remain in Portuguese. The transcription only
normalizes line breaks, list separators and minor punctuation.
`dictionary_infosiga(source = "official")` opens the source website.

A [Brazilian Portuguese version](data-dictionary-pt.html) of this article is
also available.

Three notes on reading the tables.

- **Click a row** (or the arrow) to expand the storage format, the source
  of the field, the full list of allowed values and any additional notes.
- The **Type** and **Nulls** columns describe the source schema, not necessarily
  the returned R classes. `processing = "raw"` returns every field as character;
  `"typed"` parses the documented classes; and the default `"clean"` mode adds
  missing-value, factor and flag conversions. See `?read_infosiga` for the full
  pipeline.
- In the *Source* field, **PC**, **PM** and **PRF** stand for Polícia
  Civil, Polícia Militar and Polícia Rodoviária Federal; **DETRAN-SP**
  marks fields derived by the Infosiga system itself.

```{r setup}
library(reactable)
library(htmltools)
```

```{r data}
read_dict <- function(file) {
  readr::read_tsv(
    file,
    col_types = readr::cols(.default = readr::col_character()),
    quote = "",
    na = character()
  )
}

dict_sinistros <- read_dict("dictionary-sinistros.tsv")
dict_pessoas <- read_dict("dictionary-pessoas.tsv")
dict_veiculos <- read_dict("dictionary-veiculos.tsv")
```

```{r helpers}
# Renders one dictionary as a searchable reactable. The main view shows the
# variable, description, type and nullability; format, source, allowed
# values and notes live in an expandable details row.
dict_table <- function(dict) {
  detail_row <- function(label, content) {
    div(
      class = "dict-detail-row",
      span(class = "dict-detail-label", label),
      content
    )
  }

  reactable(
    dict,
    searchable = TRUE,
    highlight = TRUE,
    defaultPageSize = 15,
    showPageSizeOptions = TRUE,
    pageSizeOptions = c(15, 30, 50),
    onClick = "expand",
    rowStyle = list(cursor = "pointer"),
    defaultColDef = colDef(vAlign = "top", headerVAlign = "bottom"),
    columns = list(
      variavel = colDef(
        name = "Variable",
        minWidth = 190,
        cell = function(value) tags$code(class = "dict-var", value)
      ),
      nome = colDef(show = FALSE),
      descricao = colDef(
        name = "Description",
        minWidth = 320,
        cell = function(value, index) {
          div(
            div(class = "dict-name", dict$nome[index]),
            div(class = "dict-desc", value)
          )
        }
      ),
      tipo = colDef(
        name = "Type",
        width = 100,
        cell = function(value) {
          if (value == "-") {
            return(span(class = "dict-type-none", "\u2014"))
          }
          span(class = paste0("dict-type dict-type-", value), value)
        }
      ),
      nulos = colDef(
        name = "Nulls",
        width = 80,
        align = "center",
        cell = function(value) {
          if (value == "sim") {
            span(class = "dict-null dict-null-yes", "yes")
          } else {
            span(class = "dict-null dict-null-no", "no")
          }
        }
      ),
      formato = colDef(show = FALSE),
      fonte = colDef(show = FALSE),
      valores = colDef(show = FALSE),
      observacao = colDef(show = FALSE)
    ),
    details = function(index) {
      row <- dict[index, ]
      values <- strsplit(row$valores, "|", fixed = TRUE)[[1]]
      div(
        class = "dict-detail",
        detail_row("Format", span(row$formato)),
        detail_row("Source", span(row$fonte)),
        if (nzchar(row$valores)) {
          detail_row(
            "Allowed values",
            if (length(values) > 1) {
              div(
                class = "dict-chips",
                lapply(values, function(v) span(class = "dict-chip", v))
              )
            } else {
              span(row$valores)
            }
          )
        },
        if (nzchar(row$observacao)) {
          detail_row("Notes", span(row$observacao))
        }
      )
    },
    theme = reactableTheme(
      borderColor = "#e4e8ee",
      highlightColor = "#f3f6fb",
      headerStyle = list(
        borderBottom = "2px solid #d7dde6",
        fontSize = "0.78rem",
        textTransform = "uppercase",
        letterSpacing = "0.04em",
        color = "#5b6675"
      ),
      searchInputStyle = list(width = "100%", maxWidth = "24rem")
    ),
    language = reactableLang(searchPlaceholder = "Search variables...")
  )
}
```

## Sinistros (crash records)

One row per confirmed crash or notification, keyed by `id_sinistro`
(`r nrow(dict_sinistros)` variables).

```{r sinistros}
dict_table(dict_sinistros)
```

## Pessoas (victims)

One row per person involved in a crash, keyed by `id_pessoa` and linked to
the other datasets by `id_sinistro` and `id_veiculo`
(`r nrow(dict_pessoas)` variables).

```{r pessoas}
dict_table(dict_pessoas)
```

## Veículos (vehicles)

One row per vehicle involved in a crash, identified by `id_sinistro` and
`id_veiculo` (`r nrow(dict_veiculos)` variables).

```{r veiculos}
dict_table(dict_veiculos)
```

## Coverage and analytical caveats

The source is published as a continuous series from 2015 onward, but its scope
has changed. These features come from the published data rather than the import
process.

- **Data from 2015--2018 cover fatal crashes only.** Non-fatal records begin in
  2019. For trends that include earlier years, restrict `sinistros` to
  `tipo_registro == "SINISTRO FATAL"` or `pessoas` to
  `gravidade_lesao == "FATAL"`. Otherwise, start the series in 2019.
- **`tipo_registro` includes notifications and confirmed crashes.** A
  `"NOTIFICACAO"` is a reported event that has not been confirmed as a crash.
  Filter this field according to the event definition required by the analysis.
- **Recent months are provisional.** DETRAN-SP reclassifies records as it
  validates them, and the final months of a release may be incomplete. Exclude
  the latest observations when comparing recent and settled periods.
- **`tempo_sinistro_obito` uses a 30-day limit.** Fatality counts therefore
  follow the 30-day convention for attributing a death to a crash.
- **Vehicle totals can differ across tables.** The sum of the `qtd_*` vehicle
  fields does not always match the number of corresponding rows in `veiculos`.
  Choose one measure and report its definition.
- **Coordinate coverage varies by year.** Spatial subsets are not a uniform
  sample over time, especially in the earliest and latest years.
- **Some crashes have no matching person or vehicle rows.** Use a left join
  from `sinistros` when the analysis must retain every crash.

These patterns describe the 2026 releases. Their frequency may change when
DETRAN-SP revises the data.

## Known inconsistencies in the source

The official PDFs contain a few internal inconsistencies. In the `pessoas`
dictionary, the format for `ano_mes_sinistro` and `ano_mes_obito` is printed as
`mm/aaaa`, but the permitted ranges and published data use `aaaa/mm`. Some year
ranges say "greater than 2022" although the corresponding files include 2022.
The observations for `ano_fab` and `ano_modelo` also print their inequalities
in reverse. The tables above preserve the official wording. The package's type
specifications follow the values found in the data files; identifiers and
`numero_logradouro` remain character vectors to avoid losing information.

## Source

DETRAN-SP, *Dicionário de dados* v1.5 (2026-06-16), distributed with the
INFOSIGA-SP open data at <https://infosiga.detran.sp.gov.br/>.
`dictionary_infosiga(source = "official")` opens the source website.

```{r dictionary, echo = TRUE, eval = FALSE}
infosigasp::dictionary_infosiga(source = "official")
```
